---
title: "bsc_reinforcement_excluded"
author: "Shoshana Rapley"
date: "2025-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Keeping the analyses that I did but decided not to include.

# Time over the fence

I suspect the flightless reinforcers acted as an anchor to the residents, most of whom had regained flight. How many days (night technically) over the fence?

```{r}
# read in night data
data_night <- read.csv("data/data_clean_night.csv") %>%
  # format time as posixct
  mutate(time_local = as.POSIXct(time_local),
         date_bird = as_date(date_bird)) %>%
  # add cohort info
  left_join(select(metadata, c(id, cohort))) %>%
  # keep only residents
  filter(cohort == "Resident") %>%
  # pre and post release period dates
  filter(date_bird %in% dates)

# convert to spatial points
points <- data_night %>%
  vect(geom = c("longitude", "latitude"), crs = "EPSG:4326")

# Find points outside fence
outside <- points[!relate(points, mtr, "intersects")] %>%
  as.data.frame(geom = "XY")
  
# Plot points outside fence 
ggmap(map_z14)+
  geom_point(data=outside, aes(x, y, colour = id), alpha = 0.7, size = 3)+
  geom_spatvector(data=mtr, inherit.aes = FALSE, colour = "white", fill = NA)+
  scale_colour_viridis_d()+
  theme_void()

# Days with points outside
excursion <- outside %>%
  group_by(id, date_bird) %>%
  summarise(count = length(id)) %>%
  # filter days <10 points unlikely to be an excursion
  filter(count >= 10)
```

It's really only Nutmeg and Marmalade with flight at this stage, so insufficient replication to answer this question.

## Figure

```{r}
# default theme options for all plots
theme <- function(){
  list(
    geom_line(),
    geom_ribbon(alpha = 0.2),
    xlab(element_blank()),
    scale_x_date(date_breaks = "1 month", date_labels = "%b"),
    theme_classic(),
    ggplot2::theme(axis.title.y = element_text(size = 10)),
    facet_wrap(~period, scales = "free_x")
  )
}

# list of plots
g7 <- ggplot(pred7, aes(date, predicted, ymin = lower, ymax = upper)) +
  ylab("Daily distance moved (m)") +
  theme()

g8 <- ggplot(pred8, aes(date, predicted, ymin = lower, ymax = upper)) +
  ylab("Roost displacement (m)") +
  theme()

g9 <- ggplot(pred9, aes(date, predicted, ymin = lower, ymax = upper)) +
  ylab("Distance from release (m)") +
  theme()

g10 <- ggplot(pred10, aes(date, predicted, ymin = lower, ymax = upper)) +
  ylab("Home range area (ha)") +
  theme()

# multi plot 2x2
ggarrange(g7, g8, g9, g10,
          ncol = 2, nrow = 2,
          labels = c("A", "B", "C", "D"))
```

# HMM
HMM which I think was garbage and didn't have the energy to fix. 

## Time budget HMM

Time budget can be another indicator of post-release behavioural modification. We expect them to become more efficient with their foraging over time as they learn where to find and exploit resources. So we expect to see more direct travel and less time spent foraging over time. 

How do they allocate their nocturnal movement? Hidden markov model.

```{r}
# read in cleaned night data
data_night <- read.csv("data/data_clean_night.csv") %>%
  # drop columns notneeded
  dplyr::select(!c(sunrise, sunset, suntime, tod)) %>%
  # format date time in posixct
  mutate(time_local = as.POSIXct(time_local),
         date_bird = as_date(date_bird))

# prep data for moveHMM
data_hmm <- data_night %>%
  prepData(type = "UTM", coordNames = c("easting", "northing"),
           covNames = c("acceleration_raw_x", "acceleration_raw_y", "acceleration_raw_z", "altitude")) %>%
# add vector of the dynamic body acceleration (VEDBA)
  mutate(vedba = ((sqrt(acceleration_raw_x^2 + acceleration_raw_y^2 + acceleration_raw_x^2))/1000))
```

### Test model

Create a test subset of one bird for one month (Nutmeg in June 2023). Investigate step and angle statistics to help define starting parameters. 

```{r}
# Test subset, Nutmeg 1 month
dates <- seq(as.Date("2023-06-01"), as.Date("2023-07-01"), "days")
test <- filter(data_hmm, id == "Nutmeg" & date_bird %in% dates)

# format as amt and calculate step lengths
steps <- test %>%
  # format as amt track - add columns as needed
  make_track(.x = x, .y = y, .t = time_local, id = id, date_bird = date_bird)

# check sample rate
summarize_sampling_rate(steps)

# resample track to 5 minutes
steps <- track_resample(steps, rate = minutes(5), tolerance = minutes(1))

# step statistics
hist(test$step, xlab = "step", main = "", breaks = 30)
summary(test$step)

# angle statistics
hist(test$angle, breaks = seq(-pi, pi, length = 15), xlab = "angle", main = "")
summary(test$angle)

```

#### Number of states

Compare a three state and four state model to see what performs better.

```{r}

# 3 state model parameters
stepPar <- c(
  10, 60, 200,   # Means
  5, 20, 150,    # Standard deviations 
  0.15, 0, 0.5   # Zero-mass values
)

anglePar <- c(
  0, 0, 0,      # Means
  1.5, 7, 3     # Concentrations
)

# 3 state model
hmm3 <- momentuHMM::fitHMM(data = test, 
               nbStates = 3,
               dist = list(step = "gamma", angle = "vm"),
               Par0 = list(step = stepPar, angle = anglePar),
               estAngleMean = list(angle=TRUE),
               formula = ~vedba
               )

hmm3
plot(hmm3)

# Step Parameters for 4 States
stepPar <- c(
  10, 60, 150, 300,   # Means 
  5, 15, 100, 150,    # Standard deviations 
  0.1, 0, 0.3, 0.15   # Zero-mass values 
)

# Angle Parameters for 4 States
anglePar <- c(
  0, 0, 0, 0,         # Means 
  1.5, 5, 2, 3        # Concentrations
)

# 4 state model
hmm4 <- momentuHMM::fitHMM(data = test, 
               nbStates = 4,
               dist = list(step = "gamma", angle = "vm"),
               Par0 = list(step = stepPar, angle = anglePar),
               estAngleMean = list(angle=TRUE),
               formula = ~vedba
               )

hmm4
plot(hmm4)

# compare 3 and 4 state models
AIC(hmm3, hmm4)
```

Prefer the 4 state model both visually and on AIC.

Model      AIC
1  hmm4 19401.37
2  hmm3 19443.89

#### Optimise parameters

Try a range of starting values to optimise parameters.

```{r}
# For reproducibility
set.seed(12345)

# Number of tries with different starting values
niter <- 10

# Save list of fitted models
allm <- list()

# Save list of starting parameters
parameters <- list()

for(i in 1:niter) {
  # Step length mean
  stepMean0 <- runif(4, min = c(5, 50, 100, 200), max = c(50, 150, 300, 400))
  
  # Step length standard deviation
  stepSD0 <- runif(4, min = c(5, 30, 50, 100), max = c(30, 100, 150, 200))
  
  # Zero mass
  zeroMass0 <- runif(4, min = c(0.5, 0.05, 0.01, 0.2), max = c(0.9, 0.3, 0.1, 0.3))
  
  # Turning angle mean
  angleMean0 <- c(0, 0, 0, 0)
  
  # Turning angle concentration
  angleCon0 <- runif(4, min = c(0.5, 3, 5, 3), max = c(3, 10, 15, 10))
  
  # trying another angle concentration now I understand this
  #angleCon0 <- runif(4, min = c(0.5, 0.1, 3, 5), max = c(3, 2, 10, 15))  
  # no this was worse
  
  # Parameter table
  parameters[[i]] <- data.frame(
    state = rep(1:4),
    stepMean0 = stepMean0,
    stepSD0 = stepSD0,
    zeroMass0 = zeroMass0,
    angleMean0 = angleMean0,
    angleCon0 = angleCon0
  )
  
  # Fit model
  stepPar0 <- c(stepMean0, stepSD0, zeroMass0)
  anglePar0 <- c(angleMean0, angleCon0)
  
  allm[[i]] <- tryCatch({
    momentuHMM::fitHMM(
      data = test,
      nbStates = 4,
      dist = list(step = "gamma", angle = "vm"),
      Par0 = list(step = stepPar0, angle = anglePar0),
      estAngleMean = list(angle = TRUE),
      formula = ~vedba
    )
  }, error = function(e) NULL)
}

# Extract likelihoods of fitted models
allnllk <- unlist(lapply(allm, function(m) m$mod$minimum))

# Index of best fitting model (smallest negative log-likelihood)
whichbest <- which.min(allnllk)

# Best fitting model
mbest <- allm[[whichbest]]
mbest
plot(mbest)
plotStates(mbest)

# Best starting parameters 
pbest_start <- parameters[[whichbest]]

# Best outcome parameters
pbest_outcome <- mbest$mle
```

Best outcome parameters:

$step
           state 1    state 2     state 3     state 4
mean     7.2211441 28.7419041 93.49649922 253.7447067
sd       5.6455927 23.2401216 53.89146576 312.9671872
zeromass 0.1559012  0.1669094  0.01833069   0.2101607

$angle
                state 1    state 2    state 3   state 4
mean          0.1170206 -0.1668971 0.02669791 0.2645331
concentration 1.1835577  1.3724136 2.71667136 1.9204624

$beta
               1 -> 2    1 -> 3    1 -> 4    2 -> 1     2 -> 3    2 -> 4
(Intercept) -5.204991 -8.829121 -2.213485 -6.221311 -10.254742 -4.297407
vedba        6.313020 -3.078513 -1.617744  5.398456  -4.197239  4.724587
               3 -> 1    3 -> 2     3 -> 4    4 -> 1    4 -> 2    4 -> 3
(Intercept) -8.220560 -2.225103  0.4526661 -2.324431 -0.453848  1.538068
vedba       -3.345948  1.189772 -2.1410536  1.063941  1.259101 -3.388000

$delta
               state 1   state 2      state 3      state 4
ID:Nutmeg 0.0004158271 0.9993644 5.228222e-05 0.0001674458

### Run HMM

Apply best parameters for a 4 state model to all birds.

```{r}
# Step Parameters 
stepPar <- c(
  7, 29, 93, 253,   # Means
  6, 23, 54, 313,    # Standard deviations 
  0.16, 0.17, 0.02, 0.21   # Zero-mass values 
)

# Angle Parameters for 4 States
anglePar <- c(
  0.11, -0.17, 0.03, 0.26,         # Means 
  1.18, 1.37, 2.72, 1.92        # Concentrations
)

# 4 state model (approx. 7 hours to run) ðŸ¢ðŸ¢
hmm_all <- momentuHMM::fitHMM(data = data_hmm, 
               nbStates = 4,
               dist = list(step = "gamma", angle = "vm"),
               Par0 = list(step = stepPar, angle = anglePar),
               estAngleMean = list(angle=TRUE),
               formula = ~vedba
               )

# see states
hmm_all

# plot results
plot(hmm_all)

# check residuals
plotPR(hmm_all)

# Decode states
data_hmm2 <- data_hmm %>%
  mutate(states = viterbi(hmm_all))

# save output
write.csv(data_hmm2, "data/data_hmm.csv", row.names = FALSE)

```

$step
$step$est
           state 1    state 2    state 3     state 4
mean     8.2208094 20.3895440 57.8629585 129.4745424
sd       6.6718586 17.2595917 33.1696834 142.4877739
zeromass 0.1835534  0.1435767  0.1017199   0.1056076

$step$se
             state 1     state 2     state 3    state 4
mean     0.059195569 0.158752114 0.504588470 1.62585187
sd       0.059392166 0.135826249 0.350108882 1.68770684
zeromass 0.001644487 0.001121659 0.002745764 0.00326152

$step$lower
           state 1    state 2     state 3      state 4
mean     8.1047882 20.0783956 56.87398323 126.28793127
sd       6.5554521 16.9933771 32.48348260 139.17992928
zeromass 0.1803303  0.1413783  0.09633832   0.09921516

$step$upper
           state 1    state 2    state 3     state 4
mean     8.3368306 20.7006924 58.8519337 132.6611535
sd       6.7882651 17.5258063 33.8558842 145.7956185
zeromass 0.1867766  0.1457751  0.1071015   0.1120001


$angle
$angle$est
                    state 1     state 2     state 3    state 4
mean          -0.0006350116 -0.02490599 0.005901283 0.02889899
concentration  0.9849618919  1.34078324 2.855987790 1.62310950

$angle$se
                  state 1     state 2     state 3    state 4
mean          0.008772497 0.004360951 0.005746417 0.01000978
concentration 0.009074356 0.008098515 0.040103562 0.01989011

$angle$lower
                  state 1    state 2      state 3     state 4
mean          -0.01782879 -0.0334533 -0.005361488 0.009280183
concentration  0.96717648  1.3249104  2.777386254 1.584125600

$angle$upper
                 state 1     state 2    state 3    state 4
mean          0.01655877 -0.01635868 0.01716405 0.04851779
concentration 1.00274730  1.35665604 2.93458933 1.66209339


$gamma
$gamma$est
            state 1    state 2      state 3    state 4
state 1 0.931458331 0.05761075 2.899808e-05 0.01090192
state 2 0.025798292 0.89292770 4.665911e-02 0.03461490
state 3 0.002059385 0.20857863 7.299991e-01 0.05936288
state 4 0.036632497 0.16340398 1.137106e-01 0.68625289

$gamma$se
            state 1     state 2     state 3      state 4
state 1 0.001852868 0.001856780         NaN 0.0009593681
state 2 0.001035936 0.001734979 0.001537815 0.0010952092
state 3 0.001406860 0.004757413 0.005098452 0.0034170704
state 4 0.002882481 0.006547791 0.005798381 0.0077365855

$gamma$lower
            state 1    state 2    state 3     state 4
state 1 0.927736437 0.05407762        NaN 0.009173398
state 2 0.023843831 0.88947937 0.04373596 0.032531334
state 3 0.000539143 0.19940768 0.71989062 0.053008962
state 4 0.031383508 0.15097250 0.10283144 0.670894349

$gamma$upper
            state 1    state 2    state 3    state 4
state 1 0.935001961 0.06135974        NaN 0.01295190
state 2 0.027908379 0.89628098 0.04976747 0.03682684
state 3 0.007832716 0.21805649 0.73987462 0.06642498
state 4 0.042720676 0.17664613 0.12557970 0.70121389

### Time budget

```{r}
# read in data with hmm states
data_hmm <- read.csv("data/data_hmm.csv")

# calculate proportion of fixes in each state per bird per day
budget_daily <- data_hmm %>%
  group_by(ID, date_bird) %>%
  # count of states
  count(states) %>%
  # pivot to wide format
  pivot_wider(names_from = 3, values_from = 4) %>%
  # rename state columns
  rename(state1 = 3,
         state2 = 4,
         state3 = 5,
         state4 = 6) %>%
  # replace NAs with zeros
  replace(is.na(.), 0) %>%
  # calculate proportion
  mutate(total = sum(state1, state2, state3, state4),
         state1p = state1/total,
         state2p = state2/total,
         state3p = state3/total,
         state4p = state4/total) %>%
  arrange(date_bird) %>%
  # rename ID
  rename(id = ID)

# save to file
write.csv(budget_daily, "results/daily_time_budget.csv", row.names = FALSE)

# summarise over time
budget <- budget_daily %>%
  # calculate mean and sd
  group_by(id) %>%
  summarise(s1m = mean(state1p),
            s2m = mean(state2p),
            s3m = mean(state3p),
            s4m = mean(state4p)) %>%
  pivot_longer(2:5, names_to = "state", values_to = "budget") %>%
  # add cohort
  left_join(select(metadata, c(id, cohort)))

# plot time budget mean by bird
ggplot(budget)+
  geom_bar(aes(x = id, y = budget, fill = state), position="fill", stat="identity")+
  theme_minimal()+
  scale_fill_viridis_d()+
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0, hjust = 1))+
  facet_wrap(~cohort, scales="free_x")
```

I had a follow up question which is how well vedba aligns with states - where I expect vedba to go 1 < 2 < 3 < 4

```{r}
# read in data with hmm states
data_hmm <- read.csv("data/data_hmm.csv") %>%
  mutate(states = as_factor(states))

# plot
ggplot(data_hmm)+
  geom_boxplot(aes(states, vedba, group = states))

# aov
temp <- aov(vedba ~ states, data = data_hmm)
summary(temp)

# post hoc tukey
TukeyHSD(temp)
```

VEDBA differed between all states except 1-2 comparison. But step lengths longer in state 2 than state 1.

### HMM example map

Want to include a figure with sample state movements, for a flying and non flying bird. 

```{r}
# hmm subset, Nutmeg 1 week
dates1 <- seq(as.Date("2023-04-22"), as.Date("2023-05-01"), "days")

sample_hmm1 <- read.csv("data/data_hmm.csv") %>%
  filter(ID == "Nutmeg" & date_bird %in% dates1) %>%
  mutate(states = as_factor(states)) %>%
  arrange(date)

## create segments to colour path
segments1 <- sample_hmm1 %>%
  mutate(next_lon = lead(longitude),
         next_lat = lead(latitude),
         next_state = lead(states)) %>%
  filter(!is.na(next_lon))

## plot on ggmap background
map1 <- ggmap(map_z14) +
  geom_spatvector(data = mtr, inherit.aes = FALSE, fill = NA, colour = "white",
                  linetype = "longdash") +
  geom_segment(data = segments1, 
               aes(x = longitude, y = latitude,
                   xend = next_lon, yend = next_lat,
                   color = states)) +
  geom_point(data = sample_hmm1, aes(longitude, latitude, colour = states)) +
  scale_colour_viridis_d() +
  theme_void()+
  annotation_north_arrow(location = "tr", 
                         height = unit(0.8, "cm"), width = unit(0.7, "cm"),
                         style = north_arrow_orienteering(text_col = "black"))+
  # add scale bar in just the resident facet panel
  annotation_scale(location = "br", text_col = "black", text_cex = 0.8,
                   pad_x = unit(0.1, "cm"), pad_y = unit(0.4, "cm"))+
  ggplot2::theme(plot.margin = margin(0, 5, 0, 5))

# hmm subset, Fauna 1 week
dates2 <- seq(as.Date("2023-06-07"), as.Date("2023-06-14"), "days")

sample_hmm2 <- read.csv("data/data_hmm.csv") %>%
  filter(ID == "Fauna" & date_bird %in% dates2) %>%
  mutate(states = as_factor(states)) %>%
  arrange(date)

## create segments to colour path
segments2 <- sample_hmm2 %>%
  mutate(next_lon = lead(longitude),
         next_lat = lead(latitude),
         next_state = lead(states)) %>%
  filter(!is.na(next_lon))

## plot on ggmap background
map2 <- ggmap(map_z15) +
  geom_spatvector(data = mtr, inherit.aes = FALSE, fill = NA, colour = "white",
                  linetype = "longdash") +
  geom_segment(data = segments2, 
               aes(x = longitude, y = latitude,
                   xend = next_lon, yend = next_lat,
                   color = states)) +
  geom_point(data = sample_hmm2, aes(longitude, latitude, colour = states)) +
  scale_colour_viridis_d() +
  theme_void()+
  annotation_north_arrow(location = "tr", 
                         height = unit(0.8, "cm"), width = unit(0.7, "cm"),
                         style = north_arrow_orienteering(text_col = "black"))+
  # add scale bar in just the resident facet panel
  annotation_scale(location = "br", text_col = "black", text_cex = 0.8,
                   pad_x = unit(0.1, "cm"), pad_y = unit(0.4, "cm"))+
  ggplot2::theme(plot.margin = margin(0, 5, 0, 5))

# multi plot
ggarrange(map1, map2,
          ncol = 2, nrow = 1,
          labels = c("A", "B"),
          common.legend = TRUE, legend = "bottom")
```

### m5: Time budget

Does time allocated to foraging differ between the social groups?

```{r}
# read in data and filter to reinforcement period
datam5 <- read.csv("results/daily_time_budget.csv") %>%
  # add metadata
  left_join(dplyr::select(metadata, c("id", "cohort"))) %>%
  # format dates as dates
  mutate(date = as_date(date_bird)) %>%
  # limit to reinforcement period
  filter(date %in% rein_dates) %>%
  # add a time elapsed post-release variable
  mutate(elapsed = as.numeric(date - as_date("2023-06-10"))) %>%
  # add column for membership to social group
  mutate(social = ifelse(id %in% club, "Reinforcer - exclusive", ifelse(cohort == "Resident", "Resident", "Reinforcer - mixed"))) %>%
  # convert group to factor and set resident as the intercept
  mutate(social = factor(social, levels = c("Resident", "Reinforcer - mixed", "Reinforcer - exclusive")))

# create social group means
sum5 <- datam5 %>%
  group_by(date, social) %>%
  summarise(dist = mean(state2p),
            upper = dist + 1.96 * std.error(state2p),
            lower = dist - 1.96 * std.error(state2p)) %>%
  arrange(date)

# overall means
mean5 <- datam5 %>%
  group_by(social) %>%
  summarise(mean = mean(state2p),
            se = std.error(state2p))

# and pivot longer for barchart
datam5bar <- datam5 %>%
  select(!3:7) %>%
  pivot_longer(3:6, names_to = "state", values_to = "proportion")

# plot time budget by social group
ggplot(datam5bar)+
  geom_bar(aes(x = social, y = proportion, fill = state),
           position="fill", stat="identity")+
  xlab("Social group")+ 
  ylab("Time budget allocation")+
  theme_minimal()+
  scale_fill_viridis_d()

# plot foraging allocation over time as social group means
ggplot(sum5, aes(date, dist, ymin = lower, ymax = upper)) +
  geom_path(aes(colour = social)) +
  geom_ribbon(aes(fill = social), alpha = 0.2) +
  xlab("Days post-reinforcement") + 
  ylab("Proportion time allocated to foraging %") +
  theme_minimal() +
  scale_fill_viridis_d()+
  scale_colour_viridis_d()

# distribution of allocation
hist(datam5$state2p)
ggqqplot(datam5$state2p)

# Adjust values slightly so that 0 < x < 1
datam5 <- datam5 %>%
  mutate(adjusted = (state2p + 0.001) / 1.002)

hist(datam5$adjusted)
ggqqplot(datam5$adjusted)

# test difference between the groups with interaction of time as a quadratic
m5 <- glmmTMB(adjusted ~ social * (scale(elapsed) + I(scale(elapsed)^2)) + (1|id), 
              data = datam5,
              family = beta_family(link = "logit"),
              na.action = "na.fail")

# model selection
dm5 <- dredge(m5)
dm5

# parsimonious model where delta <2 and fewest predictors 
# in this case the second model had delta AICc of 1 and 2 fewer predictors
best5 <- get.models(dm5, delta<2)[[2]]

# check model
check_model(best5)

# summary
summary(best5)

# post-hoc comparison of means
emmeans(best5, pairwise ~ social, type = "response")

# post-hoc comparison of slopes
emtrends(best5, pairwise ~ social, var = "elapsed")

# plot effects
plot_model(best5)

# marginal effects 
pred5 <- predict_response(best5, terms = c("elapsed [all]", "social"),
                          bias_correction = TRUE)

# plot predictions
ggplot(pred5, aes(x, predicted, fill = group, ymin = conf.low, ymax = conf.high)) +
  geom_line(aes(colour = group))+
  geom_ribbon(alpha = 0.2)+
  xlab("Days post-reinforcement") + 
  ylab("Predicted values of time allocated to foraging %") +
  theme_minimal()+
  scale_fill_viridis_d()+
  scale_colour_viridis_d()
```

**Results:**
The best fitting model included social group and time (quadratic only), the interaction between social group and the quadratic, and individual as a random effect.

The reinforcer groups (both mixed and exclusive) are more likely to exhibit higher foraging allocation (mean 64% for mixed and 67% for exclusive) than the residents (mean 47%, p < 0.0001). The reinforcers did not differ from one another. 

Linear time was not included in the model - instead a significant positive quadratic effect.The quadratic effect was significantly less for reinforcers than residents (p < 0.034). 

**Interpretation:**
Both reinforcers have higher allocations than the residents, probably because residents have overall better foraging efficiency. 

The quadratic trend may be associated with seasonal food availability.

